---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```




<!-- badges: start -->
[![R-CMD-check](https://github.com/mdsumner/sooty/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mdsumner/sooty/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of sooty is to read data of interest to Southern Ocean research. 

There are data sets like the 25km resolution south polar stereographic sea ice concentration, available via `readnsidc25kmS()`: 

```{r south}
library(sooty)
readnsidc25kmS()
```

You can immediately see what range of dates is available by setting `latest = FALSE`: 


```{r oldest}
readnsidc25kmS(latest = FALSE)
```

And, explore exactly what source files are available with the full table: 

```{r files}
files <- nsidc25kmSfiles()

range(files$date)
range(diff(files$date))  ## there are some gaps, it's every two days to start and some are missing

diff(range(files$date))  ## the number of potential data days

nrow(files)  ## the actual number of data days
```

This is a very experimental begin at replacing [raadtools](https://github.com/AustralianAntarcticDivision/raadtools) with a package anyone can use. 

Apart from north and south 25km sea ice we also have higher resolution sea ice data for the southern hemisphere. 

```{r amsr}
readamsr()

```

### Global SST 0.25 degree from 1982-current

```{r sst}
readoisst()
```



### Note for MacOS users (sadly)

On MacOS the best GDAL we can get is 3.5.3 which is sadly too old for some of these NetCDF files (these can be worked around but I don't want to do that here). 

At your own risk, we've had success installing GDAL latest. 

```
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)" && brew install gdal --HEAD
```


We haven't yet figured out what is the minimum version needed for this package to work well with all the datasets, or how to get a particular commit/tag/version (WIP). 


## Code of Conduct
  
Please note that the idt project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

